---
title: "R Notebook"
output: html_notebook
---

### LAB 1: Bulk download main packages {- .hidden .unlisted}

Now we are going to download most of the packages you need for these
labs so that you do not need to do this each week. We're doing it in
groups of 5 to make it easier to check for errors.

**Copy each of these commands into the R-CONSOLE and press enter to
run.**

A load of text should stream down your console. Just let it run until
the console is back to the \> symbol and you can type. It might take
several minutes. Sometimes this will give you an error and not run,

**WHAT DOES SUCCESS LOOK LIKE?**

If it works, you should see some text like "the downloaded binary
packages are in....". If not, there will be errors.

```{r, warnings=FALSE, message=FALSE, eval=FALSE}

install.packages(c("tidyverse","abind","car","corrplot","dismo"))
install.packages(c("elevatr","ggpubr","ggstatsplot","hrbrthemesv")) 
install.packages(c("IMTest","ISLR", "kableExtra","lattice","matlab"))
install.packages(c("maps","mapdata","maptools","mapview","ncdf4","skimr"))
install.packages(c("olsrr", "plotly","RColorBrewer"))
install.packages(c("rmdformats", "stargazer","units","viridis","yarrr","vembedr","GGally"))
install.packages(c("beeswarm","corrplot","equatiomatic"))
install.packages(c("ggExtra","ggridges","ggthemes","ztable"))
install.packages(c("Stat2Data","USAboundaries","hrbrthemes"))
install.packages(c("knitr" , "readxl", "remotes"))

```

```{r, warnings=FALSE, message=FALSE, eval=FALSE}
# The ones below are where there are often errors so we will go more slowly
install.packages(c("rgdal","rgeos","rmapshaper"))
install.packages(c("sf","sp","raster","terra"))
install.packages(c("tidycensus","tigris","tmap"))

```

Sometimes you need to install a package from a different place, like
github. To do this we use a more complex command from the remotes
library. Sometimes when doing so, R will ask questions IN THE CONSOLE.
It will wait for your answer until you type it and press enter! Choose
to install all sub-packages.

```{r, warnings=FALSE, message=FALSE, eval=FALSE}
# Sometimes you need to install a package from a different place, like github. 
# To do this we use a more complex command from the remotes library
install.packages(c("knitr" , "readxl", "remotes"))
remotes::install_github("ropensci/USAboundariesData")
remotes::install_github("ropenscilabs/rnaturalearthhires")
remotes::install_github("isciences/exactextractr")

```
